---
title: "Controlling for other variables"
author: "Michał Bojanowski"
date: "July 22, 2016"
abstract: "This is a short non-technical note about what does it mean to control for other variables in statistical multivariate analysis."
output: 
  html_document:
    number_sections: yes
    highlight: pygments
---


---

```{r setup, include=FALSE, cache=FALSE}
library(dplyr)
library(broom)
library(tidyr)
library(knitr)

knitr::opts_chunk$set(
  echo = FALSE,
  cache = TRUE
  )

set.seed(0)
```


Questions:

- What does it mean to "control for other variables" in multivariate analysis?
- What does it mean that "an effect of variable $X$ is, *ceteris paribus* equal to $\beta_x$"?


As the discussion is about *effects* of variables, it is necessary to state what an *effect of $X$ on $Y$* is.

> **Effect** of variable $X$ on variable $Y$ is an expected difference in $Y$, say $\Delta Y$, corresponding to a difference in $X$, say $\Delta X$. In other words, the efffect is $\frac{\Delta Y}{\Delta X}$.

In multiplicative or log-linear models (e.g. logistic regression, Cox regression, etc.) the effect of $X$ is usually expressed in odds-ratios. We then refer to expected multiplicative change in odds of $Y$ as a response to a change in $X$.

For example:

- *An effect of education (measured in years of schooling) on annual income (measured in dollars) is 1000* -- Given a pair of persons in which one person has one more year of schooling than the other, the more educated person is expected on average to have an income higher by $1000 than the other person.
- *The [multiplicative] effect of past experience of unemployment on risk of future unemployment is 2* -- Given a pair of persons in which one person has an episode of unemployment in the past, the odds/risk for experiencing unemployment again are two-times higher for the person with the past unemployment experience.









# Reality

We might think of variables connected causally in the following way:

```{r data}
mod <- "
  y ~ 1*x1 + 2*x2 + 3*x3 + 4*x4
  x3 ~ 1.2 * x2
  x1 ~~ 0*x4
  x2 ~~ 0*x4
  x1 ~~ 0.2*x2
"
d <- lavaan::simulateData(mod, sample.nobs = 1000) %>%
  select(x1, x2, x3, x4, y) %>%
  as_data_frame()
fit <- lavaan::sem(mod, data=d, fixed.x=FALSE)
semPlot::semPaths(fit, "est", fixedStyle = 1)
```

- All x-variables cause $y$.
- Variable x2 also causes x3.
- Variables x1 and x2 might be correlated due to common causes that do not interest us.
- Variable x4 is *not* correlated with x1, x2, and x3.

If we collect the data, we might endup with a dataset with `r ncol(d)` columns for variables 
`r paste(names(d), collapse=", ")` measured on `r nrow(d)` subjects.
First few rows of such dataset might look like this:

```{r}
d %>% head() %>% kable()
```

And the summary statistics of the variables are the following:

```{r}
d %>%
  summarise_each(
    funs(mean, sd)
  ) %>%
  gather(what, value) %>%
  separate(what, c("Variable", "f")) %>%
  spread(f, value) %>%
  kable()
```



Variables are correlated, which can be assessed with a correlation matrix:

```{r}
d %>% cor() %>% kable()
```

... or a matrix of scatterplots:

```{r}
pairs(d)
```


```{r}
by_cyl <- group_by(mtcars, cyl)
do(by_cyl, head(., 2))

models <- by_cyl %>% do(mod = lm(mpg ~ disp, data = .))
models

summarise(models, rsq = summary(mod)$r.squared)
models %>% do(data.frame(coef = coef(.$mod)))
models %>% do(data.frame(
  var = names(coef(.$mod)),
  coef(summary(.$mod)))
)

models <- by_cyl %>% do(
  mod_linear = lm(mpg ~ disp, data = .),
  mod_quad = lm(mpg ~ poly(disp, 2), data = .)
)
models
compare <- models %>% do(aov = anova(.$mod_linear, .$mod_quad))
# compare %>% summarise(p.value = aov$`Pr(>F)`)

```


    
    
# Ceteris Paribus

```{r models}
flist <- list(
  full = f <- formula(y ~ x1 + x2 + x3 + x4),
  ry = ry <- update(f, . ~ . - x3),
  rx3 = update(ry, x3 ~ . ),
  bivariate = formula( y ~ x3 )
)

modlist <- lapply(flist, lm, data=d)

d %>%
  mutate(
    ry = resid(modlist$ry),
    rx3 = resid(modlist$rx3)
  ) -> d

modlist$x3 <- lm(ry ~ rx3, data=d)
```


## The difference between marginal and partial effect

```{r}
mod_full <- lm( y ~ x1 + x2 + x3 + x4, data=d)
mod_marginal <- lm(y ~ x3, data=d)
```




## Partial effect as an effect in bivariate regression of residuals on residuals



```{r}
mod_full <- lm( y ~ x1 + x2 + x3 + x4, data=d)
m1 <- update(m0, . ~ . - x3)
m2 <- update(m1, x3 ~ .)

d %>% mutate(
  r_y = resid(m1),
  r_x3 = resid(m2)
) -> d

m3 <- update(m0, r_y ~ r_x3)

summary(m0)
summary(m3)

all.equal( coef(m0)["x3"], coef(m3)["r_x3"], check.attributes = FALSE)
```


