---
title: "Deduplication workflow"
author: "Michał Bojanowski"
date: "May 18, 2016"
output: html_document
---

```{r setup, include=FALSE}
library(dplyr)
library(RPostgres)

knitr::opts_chunk$set(
  echo = TRUE
  )
```

# Goal

Come up with a workflow for smie-automated data deduplication.

Input:

- Data frame that needs deduplication.

Output:

- Deduplicated dataset




# Data for testing purposes

Small subset of bibliographical data from PBN:

```{r pbn, cache=TRUE}
machine <- Sys.info()["nodename"]
if( machine == "starfish") {
  pbn <- src_postgres("pbn")
} else {
  pbn <- src_postgres("pbn", host="127.0.0.1", port="15432", 
                      user="mbojan")
}

set.seed(123)

# Works
pbn %>% tbl("final_work") %>% summarise(n=n()) %>%
  collect() %>% as.numeric() -> n_works
i <- sample(n_works, 100)
pbn %>% tbl("final_work") %>% 
  filter( row_number(id_work) %in% i ) %>%
  select(id_work, dtype, original_title) -> works

# Contributions
works %>%
  left_join(tbl(pbn, "final_contribution"), by=c("id_work"="fk_work")) %>%
  select(id_contribution, fk_work, given_names, family_name) -> contributions

# 

# Get it
works <- collect(works)
contributions <- collect(contributions)
```

We have data on publications and authors from PBN stored with two tables:

- `works` is a table with publications (`r paste(names(works), collapse=", ")`)
- `contributions` is a table with "authorships" with ID of work and names of authors
  (`r paste(names(contributions), sep=", ")`).




