---
title: "Data metadata"
output: 
  html_document:
    highlight: pygments
    toc: yes
---

```{r setup, include=FALSE}
library(sjmisc)

set.seed(0)
```


# Packages

Packages/functions importing data with metadata

- haven
- `foreign::read.spss`

Packages handling data metadata

- sjmisc
- labelled



# Types of metadata

Types and examples of data metadata:

- Dataset-level
    + Description
- Variable-level
    + Variable label: textual label of a variable.
    + Unit: units in which the variable is measured
- Variable value-level
    + Value labels: textual labels of individual values.
    + Special codes (or "missing data"): variable values which have some special meaning.


# Design goals:

Loose:

- Should support all basic types: numeric, logical, character.
- Converting to factors (as in sjmisc)
- Converting from factors (as in sjmisc)



# Syntax musings

Variable labels. Like in sjmisc:

```{r eval=FALSE}
x <- 1:5

y <- set_label(x, "This is a label")
set_label(x) <- "This is a label"

new_df <- set_label()
set_label(df) <- "This is a label applied to all columns of df"
```





# Generic setting/getting attributes

```{r}
set_varlevel <- function(x, ...) UseMethod("set_varlevel")

set_varlevel.default <- function(x, value, attr.name) {
  stopifnot(is.atomic(x))
  attr(x, attr.name) <- value
  x
}

set_varlevel.data.frame <- function(x, value, attr.name) {
  len <- length(value)
  if( len == 1) {
    # Same value for all
    rval <- lapply(x, function(v) {
      attr(v, attr.name) <- value
      v
    } )
    return(rval)
  } else {
    # Values must be supplied for all
    stopifnot( len == ncol(x) )
    for( i in seq(along=names(x)) ) {
      attr(x[[i]], attr.name) <- value[i]
    }
    return(x)
  }
}
```










# Artificial data

```{r data}
N <- 20
d <- data.frame(
  num = sample(1:4, N, replace=TRUE),
  ch = letters[sample(1:4, N, replace=TRUE)],
  lg = as.logical(sample(0:1, N, replace=TRUE))
)
d
```













# Setting labels

Function `set_label` sets "variable label", function `set_labels` sets "value labels".

## Individual vectors

Numeric vector

```{r numeric_vector}
x <- d$num
set_label(x) <- "This is variable label"
set_labels(x) <- c("jeden"=1)
x
```

Character vector

```{r character_vector}
x <- d$ch
set_label(x) <- "This is variable label"
set_labels(x) <- c("aaaa"="a")
x
```

Logical vector

```{r logical_vector}
x <- d$lg
set_label(x) <- "This is variable label"
set_labels(x) <- c("no"=FALSE)
x
```


## Data frames

Adding multiple variable labels in a single step

```{r data_frame}
set_label(d) <- c("Numeryczny", "tekstowy", "logiczny")
str(d)
```

Getting labels

```{r}
get_label(d)
```




# Missing values

... and other types of special codes


```{r}
x <- c(1,2,3,4,3,2,1)
set_label(x) <- "Some numeric vector"
set_labels(x) <- c("one"=1, "two"=2, "three"=3, "four"=4)
x
set_na(x, c(2,4), as.attr = TRUE)
```

